Dutch, PPs, ʀ-pronouns

### 2019-12-23 paper.tex
maandagavond
halverwege laatste sectie, conclusie moet ook nog

\subsection{\tsc{R}-pronouns as default}

@@

Important is in the insertion process is the Superset Principle.

This ensures \ref{ex:entrymolu} is not inserted into the structure in \ref{ex:structurenegmolu}, as \tsc{caus} is contained in the syntactic structure but not in the structure in the lexicon in \ref{ex:entrymolu}.

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities between the two forms.

\begin{table}[ht]
	\center
	\caption {\tit{met 't} and \tit{ermee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \tsc{wh}          & \tsc{deix}                         & \tsc{f}4   & \tsc{f}3 & \tsc{f}2  & \tsc{f}1                 & \tsc{thing}                 \\
		\midrule
    \multirow{2}{*}{w}  & \multirow{2}{*}{a}               & \tit{met}  & \multicolumn{4}{c}{\cellcolor{Gray}\tit{(ə)t}}                                \\
                      &                                    & \multicolumn{4}{c}{\tit{mee}}                                & \cellcolor{Gray}\tit{(ə)r}  \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

The \tsc{r}-pronoun I focus on is the relative pronoun \tit{waarmee} `with what', which is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with when it is combined with an \tsc{r}-pronoun.

What we do not capture is:
formal similarities between waarmee and met wat. I will talk about w, a. What I will not talk about it: mee and met
why we see the locative. I will talk about that.

What I account for: w, a, why we see the locative

What I don't: similarity between \tit{met} and \tit{mee}, that preps do not change form

and i need to say something about the assumption i make for pronouns

I do not care whether \tit{w-} and \tit{-a-} are merged before or after.

WHY DOES MEE MOVE? I DO NOT KNOW.
BUT IT DOES.
AND SO DOES WAAR.
waar maybe because it is remnant movement?

start with \tit{met wat}.

--here taking xxx apart--

\tit{'t} `it' from \ref{ex:entryt} can be used as subject, direct object and indirect object.

\ex.
\ag. 't Staat in de hal.\\
 \tsc{3sg.n.nom} stands in the hallway\\
 `It is standing in the hallway.'
\bg. Ik zie 't.\\
 I see \tsc{3sg.n.acc}\\
 `I see it.'
\bg. Ik heb 't een klap gegeven.\\
 I have \tsc{3sg.n.dat} a hit given\\
 `I hit it.'

 Pronouns in other genders alternate between nominative (non-oblique) and accusative/dative (oblique) in these contexts.

 \ex.
 \ag. Hij staat in de hal.\\
  \tsc{3sg.m.nom} stands in the hallway\\
  `He is standing in the hallway.'
 \bg. Ik zie hem.\\
  I see \tsc{3sg.m.acc}\\
  `I see it.'
 \bg. Ik heb hem een klap gegeven.\\
  I have \tsc{3sg.m.acc} a hit given\\
  `I hit him.'

@@

\tit{Er} can be used as a locative.

@@

@@

\tit{Met} can be used for as instrumental or comitative.

\exg. Ik dans met hem.\\
 I dance with him\\
 `I am dancing with him.'

Either way, it combines with accusative/dative (oblique), which can be seen on the pronouns. But only for the comitative, because for the instrumental we are getting an ʀ-pronoun. Full DPs do not show any marking, leading me to postulate the zero marker up to the dative.

\ex.
\ag. Ik dans met de man.\\
 I dance with the man\\
 `I am dancing with the man.'
\bg. Ik schilder met een kwast.\\
 I paint with een brush\\
 `I am painting with een brush.'

\ex. \begin{forest} boom
[\tsc{datP}
   [\tsc{f}3]
   [\tsc{acc}P
       [\tsc{f}2]
       [\tsc{nom}P
           [\tsc{f}1]
           [\phantom{x}]
       ]
   ]
]
{\draw (.east) node[right]{⇔ \tit{∅}}; }
\end{forest}

So \tit{met} expresses \tsc{f}4, and \tit{(ə)r} expresses \tsc{thing} and \tsc{f}1 to \tsc{f}1. \tit{(ə)r} only expresses \tsc{thing}. \tit{Ermee} and \tit{met 't} express the same features, which leaves \tsc{f}1 to \tsc{f}4 for \tit{mee}.

\begin{table}[ht]
	\center
	\caption {\tit{met 't} and \tit{ermee}}
	\begin{minipage}{0.38\linewidth}
		\begin{tabularx}{\textwidth}{ccccc}
		\toprule
    \tsc{f}4   & \tsc{f}3 & \tsc{f}2  & \tsc{f}1                 & \tsc{thing}                 \\
		\midrule
    \tit{met}  & \multicolumn{4}{c}{\cellcolor{Gray}\tit{(ə)t}}                                \\
    \multicolumn{4}{c}{\tit{mee}}                                & \cellcolor{Gray}\tit{(ə)r}  \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

@@

\tit{a}/\tit{aa} is distal, \tit{ie}/\tit{i} is proximal. Medial?

\ex.
\ag. h-ie-r\\
 here\\
 `here'
\bg. d-aa-r\\
 there\\
 `there'
\bg. d-i-t\\
 this\\
 `this'
\bg. d-a-t\\
 that\\
 `that'
\z.
\z.

WP is the interrogative/wh, whatever.

\ex. \begin{forest}
[\tsc{distP}
    [\tsc{deix3}]
    [\tsc{medP}
        [\tsc{deix2}]
        [\tsc{deix1}]
    ]
]
{\draw (.east) node[right]{⇔ \tit{a}}; }
\end{forest}\label{ex:entrya}
\z.

\section{Derivation}

@@

@@

\ex. Cyclic Override\\
Lexicalisation at a node XP overrides any previous match at a phrase contained in XP.

\ex. Backtracking\\
When spellout fails, go back to the previous cycle, and try the next option for that cycle.

@@

\section{Connecting \tit{wat} to \tit{waar} and \tit{met} to \tit{mee}}

So, \tit{waarmee} and \tit{met wat} realize the same features but their spellouts differ.

\ex.
\ag. Ik heb gekocht [waar jij mee schildert].\\
 I have bought where you with paint\\
 `I bought what you are painting with.'
\bg. Ik schilder met [wat jij hebt gekocht].\\
 I paint with what you have bought\\
 `I paint with what you bought.'

A few things should be noted here. First, position of the elements: the regular pronoun follows the adposition and the ʀ-pronoun precedes the preposition. Second, the shape of the wh-pronoun: the ʀ-pronoun differs from the non-ʀ-pronoun in the lengthening of the vowel that the final consonant is \tit{r}. Third, the shape of the adposition: \tit{met} and \tit{tot} change in \tit{mee} and \tit{toe} when they combine with ʀ-pronouns.

Now let's look at it like this: \tit{waarmee} realizes particular features. Say \tit{waar} expresses A and B and \tit{mee} expresses C and D. Say, A B C and D. \tit{met wat} realizes the same features. Saying that \tit{wat} also realizes A and B and \tit{met} also C and D misses the point of having different realizations. So instead, say \tit{wat} only realizes A and \tit{met} realizes B, C and D. Or \tit{wat} realizes A, B and C and \tit{met} only realizes D. In both cases all features are expresses, but they are distributed differently, allowing for different realizations.
\footnote{I did not address the ordering difference of the wh-element and the adposition. A topic related to (but not relevant for) this paper is the different positioning of identical adpositions in Dutch. In \ref{ex:dutchin}, \tit{in} changes meaning dependent on whether it procedes or follows the DP, it is respectively locational or directional.

@@

The different positioning of the adpositions in \tsc{r}-pronouns is driven by spellout and is, therefore, meaningless.}

I will argue for the last option, that \tit{met} spells out fewer features than \tit{mee}. The main argument for that comes from the fact that Dutch is a non-preposition stranding language, so prepositions cannot be moved out of the PP. Also, you cannot move the complement of a P to its spec, but you can subextract from them (Abels). And I show that ʀ-pronouns behave like clitic in how high they move.

In a nutshell, if all features are merged at once, we will always get an ʀ-pronoun. Due to spellout driven movement the ʀ-pronoun strands the features to be realizes by a preposition and a little more. The ʀ-pronoun is a clitic and moves high. The preposition + a little more realizes as \tit{mee} in case of \tit{waarmee}. Now if the features are not merged all at once, the regular spellout driven movements cannot take place, and we see \tit{met wat}. So, the answer of how constituency connects to change in phonological form lies in one of the core assumptions of nanosyntax: phrasal spellout spells out constituents.

\ex.
\a. \tit{waar} = w + a + r
\b. \tit{wat} = w + a + t

So the difference betwween \tit{t} and \tit{r} was actually the only difference we need to account for.

Dutch is a non P-stranding language.

\ex.
\ag. Met wie heb jij gedanst?\\
 with who have you danced\\
 `Whom did you dance with?'
\bg. *Wie heb jij gedanst met?\\
 who have you danced with\\
 `Whom did you dance with?'

Abels shows that extraction from a PP is not categorically excluded. It happens in Slavic.

--examples here--

What if ʀ-pronouns are actually clitic? And they pattern with Slavic in that they clitic can never follow Ps.

Do we have more reasons to believe that ʀ-pronouns are clitics? The non-wh ʀ-pronouns in Dutch are able to move very high, just like clitics and unlike regular pronouns in Dutch.\footnote{\ref{ex:hemaan} can also be excluded by saying that \tit{aan} and \tit{hem} cannot split.}

\ex.
\ag. Ik heb veel geld aan de badkamer uitgegeven.\\
 I have much money on the bathroom spent\\
 `I spent a lot of money on the bathroom.'
\bg. Ik heb er veel geld aan uitgegeven.\\
 I have there much money on spent\\
 `I spent a lof of money on it.'

\ex.
\ag. Ik heb veel geld aan mijn zoon uitgegeven.\\
 I have much money on my son spent\\
 `I spent a lof of money on my son'
\bg. Ik heb veel geld aan hem uitgegeven.\\
 I have much money on him spent\\
 `I spent a lot of money on him.'
\bg. *Ik heb \tbf{hem} veel geld aan uitgegeven.\\
 I have him much money on spent\\
 `I spent a lof of money on it.'\label{ex:hemaan}
\z.
\z.

This has shown that \tit{er} is small, smaller than regular pronouns. So we go for this option:

@@

Open issues:

\section{Conclusion and open issues}

@@

\ex. \begin{forest} boom
[XP
    [XP
        [X,roof]
    ]
    [YP
        [Y]
        [\phantom{x}]
    ]
]
{\draw (.east) node[right]{⇔ \tit{in}}; }
\end{forest}\label{ex:presuf}

>>>>

\subsection{\tsc{R}-pronouns as default}\label{sec:rdefault}

@@

Spellout targets only constituent \citep{starke2009}. This ensures \ref{ex:entrymolu} is not inserted into the structure in \ref{ex:structurenegmolu}, as \tsc{caus} is contained in the syntactic structure but not in the structure in the lexicon in \ref{ex:entrymolu}.

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities between the two forms. First, I identify \tit{w-} and \tit{-a-} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{ermee} `with it' and \tit{met 't} `with it'. I decompose these two expressions further, and end up arguing for a picture as shown in Table \ref{tbl:nofeatures}.
The elements \tit{w-} and \tit{-a-} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{mee} and \tit{er} together, but the distribution differs. \tit{Met} expresses less structure than \tit{mee} and \tit{'t} expresses more structure than \tit{er}. It becomes clear that the \tit{er} in fact corresponds to the locative \tit{r} in Dutch.

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \phantom{\tsc{wh}}  & \phantom{\tsc{deix}}                    & \phantom{\tsc{f}4}  & \phantom{\tsc{f}3} & \phantom{\tsc{f}2}  & \phantom{\tsc{f}1}  & \phantom{\tsc{thing}} \\
		%\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{(ə)t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{(ə)r}  \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:nofeatures}

Finally, I put everything together and show how \tit{waarmee} `with what' is inserted if all features form a constituent, and \tit{met wat} `with what' if this is not the case.

\subsection{Taking everything apart}

In this section I decompose \tit{waarmee} `with what' and \tit{met wat} `with what'. The idea is that both expessions realize the same features, and \tit{waarmee} `with what' takes precedence when all features form a constituent. The morphemes I distinguish in \tit{waarmee} `with what' are \tit{w-}, \tit{-aa-}, \tit{-r} and \tit{-mee}. Within \tit{met wat} `with what' I distinguish \tit{met}, \tit{w-}, \tit{-a-} and \tit{-t}. \ref{ex:decompose} shows this as well.

\ex.\label{ex:decompose}
\a. w -aa -r -mee
\b. met w -a -t

Let me start with the morphemes \tit{w-} and \tit{-a-} that appear in both expressions. Exactly because they appear in both expression, I do not pay much attention to them. For \tit{w-} I follow \citet{hachem2015} in that \tit{w-} realizes a WP.

The morpheme \tit{-a-} expresses deixis. Dutch distinguishes between proximal by using \tit{-ie-/-i-} and distal by using \tit{-aa-/-a-}, illustrated in \ref{ex:deixis}. I assume that lengthening or shortening of the vowel is a result of the final \tit{r}.

\ex.\label{ex:deixis}
\ag. h-ie-r\\
 here\\
\bg. d-aa-r\\
 there\\
\bg. d-i-t\\
 this\\
\bg. d-a-t\\
 that\\
 \z.

In combination with wh-elements, the distal marker \tit{-a-} is used. For the purpose of this paper I simply let \tit{-a-} correspond to \tsc{deixP}.

\ex. \begin{forest}
[deixP
    [deix, roof]
]
{\draw (.east) node[right]{⇔ \tit{a}}; }
\end{forest}\label{ex:entrya}

I put \tit{w-} and \tit{-a-} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{er-mee} `with it' and \tit{met 't} `with it'.\footnote{\tit{'rmee} `with it' is often precdeded by an /ε/, which I assume to be for phonological reasons.}

\ex.
\ag. (e)r -mee\\
there with\\
\b. met 't\\
with it\\

I repeat the examples from Section \ref{sec:rdefault} below that show that \tit{ermee} `with it' is used when it forms a proper constituent and \tit{met 't} `with it' is ungrammatical.

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:jarmee}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}

In other words, both expessions realize the same features, and \tit{ermee} `with it' takes precedence when all features form a constituent.

Let me continue with specifying the lexical entries for \tit{met 't} `with it'. \tit{'t} `it' can be used as subject, direct object and indirect object, as shown in \ref{ex:tsubobj}.

\ex.\label{ex:tsubobj}
\ag. 't Staat in de hal.\\
 \tsc{3sg.n.nom} stands in the hallway\\
 `It is standing in the hallway.'
\bg. Ik zie 't.\\
 I see \tsc{3sg.n.acc}\\
 `I see it.'
\bg. Ik heb 't een klap gegeven.\\
 I have \tsc{3sg.n.dat} a hit given\\
 `I hit it.'

Pronouns in other genders alternate between nominative (non-oblique) and accusative/dative (oblique) in these contexts, illustrated in \ref{ex:hemsubobj}.

\ex.\label{ex:hemsubobj}
\ag. Hij staat in de hal.\\
 \tsc{3sg.m.nom} stands in the hallway\\
 `He is standing in the hallway.'
\bg. Ik zie hem.\\
 I see \tsc{3sg.m.acc}\\
 `I see it.'
\bg. Ik heb hem een klap gegeven.\\
 I have \tsc{3sg.m.acc} a hit given\\
 `I hit him.'

I assume that the inanimate personal pronoun minimally contains the ontological category \tsc{thing} \citep{kayne2005}. I leave possible number and gender features out because they do not play a role in this paper. For case, I follow \citet{caha2009} that case features case features are organized the containment relation in \ref{ex:casetree}. The higher, more complex cases contain the smaller, less complex cases.

\ex. \label{ex:casetree}
\begin{forest} boom
[\tsc{comP}
    [\tsc{f5}]
    [\tsc{insP}
        [\tsc{f4}]
        [\tsc{datP}
            [\tsc{f3}]
            [\tsc{accP}
                [\tsc{f2}]
                [\tsc{nomP}
                    [\tsc{f1}]
                    [DP]
                ]
            ]
        ]
    ]
]
\end{forest}

The morpheme \tit{'t} can act as nominative, accusative and dative, as I showed in \ref{ex:tsubobj}. Taking this all together, \tit{'t} should have the lexical entry given in \ref{ex:entryt}.

@@

This lexical entry can lexicalize the \tsc{datP}, but also the \tsc{accP} and \tsc{nomP}. This is due to the Superset Princple.

In other words, a lexically stored structure does not have to be identical to the syntactic structure. It is enough if the syntactic structure is contained in the lexically stored tree.

\tit{Met} `with' can be used for as instrumental or comitative, as shown in \ref{ex:metinscom}.

\exg. Ik dans met hem.\\
 I dance with him\\
 `I am dancing with him.'\label{ex:metinscom}

Again, due to the superset principle, \tit{met} `with' can be used for comitative and as instrumental case. It combines with the dative, so the lowest feature is \tsc{f4}.

-------
now we get to \tit{er-mee}. they have to be able to lexicalize the same features as in \tit{met 't}.
-----

 \tit{Er} `there' can be used as a locative.

@@

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

@@

So \tit{met} expresses \tsc{f}4 and \tsc{f5}, and \tit{(ə)r} expresses \tsc{thing} and \tsc{f}1 to \tsc{f}3. \tit{(ə)r} only expresses \tsc{thing}. \tit{Ermee} and \tit{met 't} express the same features, which leaves \tsc{f}1 to \tsc{f}5 for \tit{mee}.

And! Nanosyntax encodes the difference between prefixes and suffixes by the difference in the shape of the lexical tree. \tit{-mee} follows \tit{er}, so I am going to call it a suffix. So:

@@

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \phantom{\tsc{wh}}  & \phantom{\tsc{deix}}                    & \phantom{\tsc{f}4}  & \phantom{\tsc{f}3} & \phantom{\tsc{f}2}  & \phantom{\tsc{f}1}  & \phantom{\tsc{thing}} \\
		%\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{(ə)t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{(ə)r}  \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:nofeatures}

% I do not care whether \tit{w-} and \tit{-a-} are merged before or after.
%
% WHY DOES MEE MOVE? I DO NOT KNOW.
% BUT IT DOES.
% AND SO DOES WAAR.
% waar maybe because it is remnant movement?
%
%
%
% Either way, it combines with accusative/dative (oblique), which can be seen on the pronouns. But only for the comitative, because for the instrumental we are getting an ʀ-pronoun. Full DPs do not show any marking, leading me to postulate the zero marker up to the dative.
%
% \ex.
% \ag. Ik dans met de man.\\
%  I dance with the man\\
%  `I am dancing with the man.'
% \bg. Ik schilder met een kwast.\\
%  I paint with een brush\\
%  `I am painting with een brush.'
%
% \ex. \begin{forest} boom
% [\tsc{datP}
%    [\tsc{f}3]
%    [\tsc{acc}P
%        [\tsc{f}2]
%        [\tsc{nom}P
%            [\tsc{f}1]
%            [\phantom{x}]
%        ]
%    ]
% ]
% {\draw (.east) node[right]{⇔ \tit{∅}}; }
% \end{forest}»

\subsection{Putting it back together}

@@

\ex. The Elsewhere Condition (\citealt{kiparsky1973}, formulated as in \citealt{caha2020}):\\
When two entries can spell out a given node, the more specific entry wins. Under the Superset Principle governed insertion, the more specific entry is the one which has fewer unused features

Spellout algorithm \citep{starke2018}

@@

\ex. Cyclic Override \citep{starke2018}:\\
Lexicalisation at a node XP overrides any previous match at a phrase contained in XP.

\ex. Backtracking \citep{starke2018}:
When spellout fails, go back to the previous cycle, and try the next option for that cycle.

\ex. Spec Formation \citep{starke2018}:\\
If Merge F has failed to spell out (even after backtracking), try to spawn a new derivation providing the feature F and merge that with the current derivation, projecting the feature F at the top node.

@@

I did not address the ordering difference of the wh-element and the adposition. A topic related to (but not relevant for) this paper is the different positioning of identical adpositions in Dutch. In \ref{ex:dutchin}, \tit{in} changes meaning dependent on whether it procedes or follows the DP, it is respectively locational or directional.

@@

The different positioning of the adpositions in \tsc{r}-pronouns is driven by spellout and is, therefore, meaningless.

\section{Conclusion and discussion}

So, the answer of how constituency connects to change in phonological form lies in one of the core assumptions of nanosyntax: phrasal spellout spells out constituents.

Just like Abels said: no preposition stranding, but extraction from the PP is ok!

@@

Open ends

@@

\ex. \begin{forest} boom
[XP
    [XP
        [X,roof]
    ]
    [YP
        [Y]
    ]
]
{\draw (.east) node[right]{⇔ \tit{in}}; }
\end{forest}\label{ex:presuf}
